Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
typo and size
tweaks

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{object_states.pdf}
  \caption{Objects states handled by the write barrier\label{fig:obj_states}}
\end{figure}

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones and a little less in the others. The reason
for that is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. STM simply
replaces the GIL in those programs. In the latter group,
the threads work on a common data structure and therefore
create much more conflicts, which limits the scalability. Here
we make active use of the STM-supported atomic blocks that
provide a scalable synchronisation mechanism.

Looking at the average overhead from switching from GIL to STM, we see
that it is \remi{$\approx 35.5\%$}. The maximum in richards is
\remi{$63\%$}.

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it rarely beats CPython's single-thread performance.  For
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

>>>>

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{object_states.pdf}
  \caption{Object states handled by the write barrier\label{fig:obj_states}}
\end{figure}

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
the others (\remi{$avg=1.9\times$} speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks in order do synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead from switching from GIL to STM, we see
that it is \remi{$\approx 35.5\%$}. The maximum in richards is
\remi{$63\%$}. In all benchmarks \emph{pypy-stm-nojit} beats
\emph{pypy-nojit} already on two threads despite of this overhead.
However, it rarely beats CPython's single-thread performance.  For
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

@@

First results promising; a lot to understand yet; a lot to improve

contributions:
\begin{itemize}
\item memory segmentation, page mapping $\rightarrow$ performance
\item STM \& GC: high allocation rate of dynlangs; shared write barrier;
  page privatisation and re-sharing by GC; used to synchronise threads;
  used to classify objects (transaction owned / not)
\item GIL replaced successfully in embarrassingly parallel benchs;
  avg overhead of $35\%$; beating GIL at 2 threads (2.5x speedup)
\item atomic blocks introduced as a scalable sync mechanism (1.9x speedup)
\item JIT needs work
\end{itemize}
